### 2023-12-04 manuscript.Rmd
Edits AB

### 1.2.2	Statistical Power for Model Comparison 
A second question that is of interest for many researchers – especially in developmental cognitive neuroscience – pertains to identifying the one theoretical model (from a set of competing models) that best explains the given data [@henson_2016; @miller_2012; @nyberg_1994]. In the case of competing models that are nested, the analytical approach to this question (and therefore also considerations on statistical power) can be addressed analogously to the procedure outlined above [see @jobst_2021 for a step-by step tutorial]. However, comparing non-nested models prohibits the use of regular $\chi^2$-based statistics and therefore poses additional methodological challenges in determining the ‘best’ model. Nevertheless, researchers should strive for a methodological approach that matches the theoretical assumptions, rather than vice versa, that is, moving away from well-grounded theoretical considerations to meet methodological constraints. In cases where analytical solutions to questions of statistical properties like statistical power are not available, randomization-based techniques offer an excellent alternative [@efron_1994]. In particular, Monte Carlo simulations have proven useful to bridging this gap [@muthen_2002, see section 2.1 for more details on Monte Carlo simulations].

@@

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model among the three competing models, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (section 3.1) and the reliability of the manifest variables (section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (section 3.3).

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (tasks 1 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

@@

### 2.1.3	Step 3: Generate 
Next, we define the mechanism to generate the synthetic data. An easy way to generating synthetic data within an SEM framework in R is the simulateData() function available in the lavaan package. In the example on memory composition we simulate continuous data, as the indicators represent accuracy values from the memory tasks. However, the simulation approach presented here can of course be extended to dichotomous or polytomous data for other use cases. Within the generate function, all parameters that are predetermined across iterations can be directly entered into the generating model syntax. However, all parameters that are supposed to be varied between different iterations of the simulation need to be accessed through the condition argument that is returned through the createDesign() function in step 2. Given our interests in the effect of loading strength and strength of covariance between the latent factors, we define these parameters in step 2 (Design) such that they can now be called via the condition argument in the generate function: 
```{r generate, echo = TRUE}
```

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as "a" for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as "b" to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). Both loading parameters a and b are stored in the list "parameters" such that they can be called via the glue command. The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The population model is defined by a model string, which specifies three latent factors (ps, pc, gen) measured by the 13 manifest indicators (task1 – task13), the residual variances as defined above and the covariance between the latent factors as defined in the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package and stored as "dat".

@@

### 2.1.4 Step 4: Analyze
In this step, we define the analyze function, which specifies how the data generated in step 3 should be analyzed. For the comparison of non-nested models, we rely on a combination of relative and absolute model fit parameters (for discussion of an alternative model selection approach, see section 4). We fit each of the competing models to each of the simulated data sets using the sem() function from the lavaan package with a maximum likelihood estimator and extract different measures of model fit using the helper function get_fitmeasures() (see funs.R in the supplementary material on GitHub), namely the Comparative Fit Index [CFI, @bentler_1990], root mean square error of approximation [RMSEA, @steiger_2016] and the Bayesian Information Criterion [BIC, @raftery_1986]. The function returns the fit measures, the simulation model that was used in the respective iteration and information on model convergence. The analyze function for the example on memory composition could thus look like this: 
```{r analyze, echo = TRUE}
```

### 2.1.5	Step 5: Summarize
In step 5, we define how the results from the analyze step should be summarized. To this end, we extract how often the model used to generate the data (step 2) was actually recovered as the best fitting model. We implemented the following decision rules for identifying the best fitting model: (i) We define cut-off criteria for an adequate model fit according to established rules as CFI > .95 and RMSEA < 0.06 [@hu_1999]. (ii) Among the models that meet the cut-off criteria, we select the best fitting model for each data set based on the relative model fit (lowest BIC value). The goal of (i) and (ii) is to check whether the ground truth model that generated the data can be identified as the best fitting model in a given simulation run. For cases in which a given dataset does not converge for one of the models, we consider the outcome of these iterations as failing to recover the correct model. We summarize the proportion of successful model recovery for each combination of model parameters in the simulation separately. The summarize function that concatenates the results across simulation iterations for the example on the structure of memory in childhood could thus look like this:
```{r summarize, echo = TRUE}
```
First, the function restructures the results from the analyze function to obtain a tidy data frame with the fit measures for each model in a separate row. This data is then summarized for each iteration. The function identifies a) the ground truth model that generated the data in this simulation iteration (sim_model) and b) the model which produced the best model fit (best). Here, the best model is defined as “0” if one of the three theoretical models did not converge, or if none of the models met the cut-off criteria as defined above. Otherwise, the best model is selected among the ones that met the cut-off criteria based on the relative model-fit.  The function returns the argument best_3, indicating how often across all iterations Model 3 was identified as the best fitting model, and best_sim, indicating how often the model that simulated the data was identified as the best fitting model (note that in the current analysis, both values are identical. best_sim can however be useful to investigate Type-I error, see section 3.3).

@@

## 3.3	How can Simulations Inform us About Type-I Error Rates?
The main part of this tutorial has focused on estimating statistical power that is required to reject a null hypothesis when it is indeed false (Type-II error). However, it is necessary to also consider the reverse — rejecting a null hypothesis that is in fact true (Type-I error). In our example, Type-I error would be committed by misidentifying Model 3 as the best fitting model when in reality, the data were generated from either Models 1 or 2. We can leverage the simulation approach to quantify Type-I error rates dependent on the different simulation conditions (e.g., inter-factor correlations, factor loadings). We conducted an analogous analysis to the one outlined above (sections 2.1.1. – 2.1.6), with the difference that Models 1 and 2 are defined as the generating population models (step 3, section 2.1.3). The outcomes of this analysis show the probability of erroneously identifying an overly complex model, when the more simplistic models generated the data (see Figure 4).

@@

Note that in this tutorial, we limited our analyses to investigating the impact of either the separability of the theoretical constructs or the impact of the reliability of the estimates by keeping the other parameter constant throughout the simulation. The interaction of these parameters may differentially impact the resultant sample size estimates. However, a thorough investigation of the interaction of the two influencing factors would exponentially increase the number of conditions in a simulation. In our example on memory component processes we therefore confined our simulation to separate analyses in the interest of conciseness. For interested readers, two additional analyses on a) the effect of loading strength on estimates of power for higher levels of inter-factor correlation (0.7 as opposed to 0.3 in the original simulation) and b) the effect of inter-factor correlation on estimates of power for lower levels of factor-loadings (0.5 as opposed to 0.7 in the original simulation) can be found as supplementary material on the github repository. Given that previous work on memory development supports the assumption of reasonably high factor loadings and moderate inter-factor correlation in the present example, these additional analyses leave the derived sample size estimates unchanged.

Further, we have presented an approach that leverages a combination of absolute and relative model fit parameters to adjudicate between competing models. While such parameter-based decisions are well established in the field, they may result in choosing one model over the other, even in cases when differences in model fit parameters are very small, as one can never conclude that two models fit equally well. An alternative approach to this criterion for model selection originates from a theory by @vuong_1989, who suggested likelihood ratio tests for non-nested models. Recent advances that build on this idea have postulated a framework that allows to test hypotheses of model distinguishability and difference in fit [ @levy_2007; @levy_2011], which can be applied to non-nested SEMs and easily implemented via new software packages in R [ @merkle_2016 ]. The resulting recommendations for the comparison of non-nested SEMs posit a stepwise procedure to test whether competing non-nested models are distinguishable in a given population and, if yes, whether one fits the data significantly better than the other. Importantly, this method derives interval estimates for differences in non-nested information criteria [ @merkle_2016 ]. Thus, this approach does not only allow to conclude that one or the other model fits the data better, but also allows for the possibility to conclude that there is insufficient evidence to determine which of two models fits a given dataset better. This stepwise procedure can be implemented as a model selection criterion in a power simulation like the one presented above. To integrate model selection via a likelihood ratio test for non-nested models into the current simulation, one would need to adapt the summarize function accordingly, such that it identifies model 3 as the best fitting model only if a) the competing models are distinguishable on the simulated data set of a given iteration, b) the Likelihood ratio test for non-nested models indicates in a significantly better fit for model 3 compared to models 1 and 2 and c) model 3 reaches the cutoffs for absolute model fit, as defined in the previous simulation. The vuongtest() function from the nonnest2 package [@merkle_2014] can be a helpful tool for integrating these tests into the simulation. While a systematic investigation of the interplay of indicator reliability, separability of latent factors and statistical power for model selection via model fit parameters versus the LRT for non-nested models is beyond the scope of this tutorial, it is worth noting that such an alternative approach may yield better model recovery in certain cases. The conditions under which model selection via the LRT for non-nested models provides better estimates of statistical power are thus an interesting avenue for future research.

>>>>

### 1.2.2	Statistical Power for Model Comparison 
A second question that is of interest for many researchers – especially in developmental cognitive neuroscience – pertains to identifying the one theoretical model (from a set of competing models) that best explains the given data [@henson_2016; @miller_2012; @nyberg_1994]. In the case of competing models that are nested, the analytical approach to this question (and therefore also considerations on statistical power) can be addressed analogously to the procedure outlined above [see @jobst_2021 for a step-by step tutorial]. However, comparing non-nested models prohibits the use of regular $\chi^2$-based statistics and therefore poses additional methodological challenges in determining the ‘best’ model. Nevertheless, researchers should strive for a methodological approach that matches the theoretical assumptions, rather than vice versa, that is, moving away from well-grounded theoretical considerations to meet methodological constraints. In cases where analytical solutions to questions of statistical properties like statistical power are not available, randomization-based techniques offer an excellent alternative [@efron_1994]. In particular, Monte Carlo simulations have proven useful to bridging this gap [@muthen_2002, see Section 2.1 for more details on Monte Carlo simulations].

@@

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in Section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model among the three competing models, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (Section 3.1) and the reliability of the manifest variables (Section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (Section 3.3).

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (ps: tasks 1 to 6, pc: tasks 5 to 9, gen: tasks 10 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

@@

### 2.1.3	Step 3: Generate 
Next, we define the mechanism to generate the synthetic data. An easy way to generating synthetic data within an SEM framework in R is the simulateData() function available in the lavaan package [@rosseel_2012]. In the example on memory composition we simulate multivariate normal data, as the indicators represent continuous values of task performance from the memory tasks. However, the simulation approach presented here can of course be extended to dichotomous or polytomous data for other use cases. Within the generate function, all parameters that are predetermined across iterations can be directly entered into the generating model syntax. However, all parameters that are supposed to be varied between different iterations of the simulation need to be accessed through the condition argument that is returned through the createDesign() function in step 2. Given our interests in the effect of loading strength and strength of covariance between the latent factors, we define these parameters in step 2 (Design) such that they can now be called via the condition argument in the generate function: 
```{r generate, echo = TRUE}
```

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as "a" for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as "b" to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). Both loading parameters a and b are stored in the list "parameters" such that they can be called via the glue_data function from the glue package [@hester_2022] to concatenate the model string. The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The population model is defined by a model string, which specifies three latent factors (ps, pc, gen) measured by the 13 manifest indicators (task1 – task13), the residual variances as defined above and the covariance between the latent factors as defined in the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package [@rosseel_2012] and stored as "dat".

@@

### 2.1.4 Step 4: Analyze
In this step, we define the analyze function, which specifies how the data generated in step 3 should be analyzed. For the comparison of non-nested models, we rely on a combination of relative and absolute model fit parameters (for discussion of an alternative model selection approach, see Section 4). We fit each of the competing models to each of the simulated data sets using the sem() function from the lavaan package [@rosseel_2012] with a maximum likelihood estimator and extract different measures of model fit using the helper function get_fitmeasures() (see funs.R in the supplementary material on GitHub), namely the Comparative Fit Index [CFI, @bentler_1990], root mean square error of approximation [RMSEA, @steiger_2016] and the Bayesian Information Criterion [BIC, @raftery_1986]. The function returns the fit measures, the simulation model that was used in the respective iteration and information on model convergence. The analyze function for the example on memory composition could thus look like this: 
```{r analyze, echo = TRUE}
```

### 2.1.5	Step 5: Summarize
In step 5, we define how the results from the analyze step should be summarized. To this end, we extract how often the model used to generate the data (step 2) was actually recovered as the best fitting model. We implemented the following decision rules for identifying the best fitting model: (i) We define cut-off criteria for an adequate model fit according to established rules as CFI > .95 and RMSEA < 0.06 [@hu_1999]. (ii) Among the models that meet the cut-off criteria, we select the best fitting model for each data set based on the relative model fit (lowest BIC value). The goal of (i) and (ii) is to check whether the ground truth model that generated the data can be identified as the best fitting model in a given simulation run. For cases in which a given dataset does not converge for one of the models, we consider the outcome of these iterations as failing to recover the correct model. We summarize the proportion of successful model recovery for each combination of model parameters in the simulation separately. The summarize function that concatenates the results across simulation iterations for the example on the structure of memory in childhood could thus look like this:
```{r summarize, echo = TRUE}
```
First, the function restructures the results from the analyze function to obtain a tidy data frame with the fit measures for each model in a separate row. This data is then summarized for each iteration. The function identifies a) the ground truth model that generated the data in this simulation iteration (sim_model) and b) the model which produced the best model fit (best). Here, the best model is defined as “0” if one of the three theoretical models did not converge, or if none of the models met the cut-off criteria as defined above. Otherwise, the best model is selected among the ones that met the cut-off criteria based on the relative model-fit.  The function returns the argument best_3, indicating how often across all iterations Model 3 was identified as the best fitting model, and best_sim, indicating how often the model that simulated the data was identified as the best fitting model (note that in the current analysis, both values are identical. best_sim can however be useful to investigate Type-I error, see Section 3.3).

@@

## 3.3	How can Simulations Inform us About Type-I Error Rates?
The main part of this tutorial has focused on estimating statistical power that is required to reject a null hypothesis when it is indeed false (Type-II error). However, it is necessary to also consider the reverse — rejecting a null hypothesis that is in fact true (Type-I error). In our example, Type-I error would be committed by misidentifying Model 3 as the best fitting model when in reality, the data were generated from either Models 1 or 2. We can leverage the simulation approach to quantify Type-I error rates dependent on the different simulation conditions (e.g., inter-factor correlations, factor loadings). We conducted an analogous analysis to the one outlined above (Sections 2.1.1. – 2.1.6), with the difference that Models 1 and 2 are defined as the generating population models (step 3, Section 2.1.3). The outcomes of this analysis show the probability of erroneously identifying an overly complex model, when the more simplistic models generated the data (see Figure 4).

@@

Note that in this tutorial, we limited our analyses to investigating the impact of either the separability of the theoretical constructs or the impact of the reliability of the estimates by keeping the other parameter constant throughout the simulation. The interaction of these parameters may differentially impact the resultant sample size estimates. However, a thorough investigation of the interaction of the two influencing factors would exponentially increase the number of conditions in a simulation. In our example on memory component processes we therefore confined our simulation to separate analyses in the interest of conciseness. For interested readers, two additional analyses on a) the effect of loading strength on estimates of power for higher levels of inter-factor correlation (0.7 as opposed to 0.3 in the original simulation) and b) the effect of inter-factor correlation on estimates of power for lower levels of factor-loadings (0.5 as opposed to 0.7 in the original simulation) can be found as supplementary material on the github repository. Given that previous work on memory development supports the assumption of reasonably high factor loadings and moderate inter-factor correlation in the present example, these additional analyses leave our conclusions unchanged.

Further, we have presented an approach that leverages a combination of absolute and relative model fit parameters to adjudicate between competing models. While such parameter-based decisions are well established in the field, they may result in choosing one model over the other, even in cases when differences in model fit parameters are very small, as one can never conclude that two models fit equally well or that there is too little data to make a decision with enough confidence. An alternative approach to this criterion for model selection originates from a theory by @vuong_1989, who suggested likelihood ratio tests for non-nested models. Recent advances that build on this idea have postulated a framework that allows to test hypotheses of model distinguishability and difference in fit [ @levy_2007; @levy_2011], which can be applied to non-nested SEMs and easily implemented via new software packages in R [ nonnest2, @merkle_2016]. The resulting recommendations for the comparison of non-nested SEMs posit a stepwise procedure to test whether competing non-nested models are distinguishable in a given population and, if yes, whether one fits the data significantly better than the other. Importantly, this method derives interval estimates for differences in non-nested information criteria [ @merkle_2016 ]. Thus, this approach does not only allow to conclude that one or the other model fits the data better, but also allows for the possibility to conclude that there is insufficient evidence to determine which of two models fits a given dataset better. This stepwise procedure can be implemented as a model selection criterion in a power simulation like the one presented above. To integrate model selection via a likelihood ratio test for non-nested models into the current simulation, one would need to adapt the summarize function accordingly, such that it identifies model 3 as the best fitting model only if a) the competing models are distinguishable on the simulated data set of a given iteration, b) the Likelihood ratio test for non-nested models indicates in a significantly better fit for model 3 compared to models 1 and 2 and c) model 3 reaches the cutoffs for absolute model fit, as defined in the previous simulation. The vuongtest() function from the nonnest2 package [@merkle_2014] can be a helpful tool for integrating these tests into the simulation. While a systematic investigation of the interplay of indicator reliability, separability of latent factors and statistical power for model selection via model fit parameters versus the LRT for non-nested models is beyond the scope of this tutorial, it is worth noting that such an alternative approach may yield better model recovery in certain cases. The conditions under which model selection via the LRT for non-nested models provides better estimates of statistical power are thus an interesting avenue for future research.
